---
title: "MB5370_M2_WS1"
author: "Kaytlin Edwards"
format: html
editor: visual
---

```{r}
#| include: false

## Inventory every active object currently residing in session RAM
objects()

## Purge global environment
rm(list = ls())

# Confirm that global session memory now completely vacant
objects()
```

# Penguin Morphology

### Load Data and Packages

```{r}
# Run the data preparation script
source("code/diw_sensor.R")
```

## Variation in Body Mass Across Species and Island

```{r}
# Calculate mean body mass by species and island
mass_summary <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass_g = mean(body_mass_g, na.rm = TRUE)
  )

mass_summary
```

## Body Mass Distribution

```{r}
# Visualise body mass across species and islands
penguins |>
   filter(!is.na(body_mass_g)) |>
  ggplot(aes(x = species, y = body_mass_g, fill = island)) +
  geom_boxplot() +
  labs(
    title = "Body Mass Distribution of Penguins by Species and Island",
    x = "Species",
    y = "Body Mass (g)",
    fill = "Island"
  ) +
  theme_minimal()
```

## Summary Table of Body Mass by Species and Island

```{r}
# Create summary table by species and island
mass_summary <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass = sd_mass / sqrt(n),
    .groups = "drop"
  )

# Display summary table
knitr::kable(
  mass_summary,
  digits = 1,
  col.names = c(
    "Species",
    "Island",
    "Mean Body Mass (g)",
    "SD (g)",
    "Sample Size",
    "SE (g)"
  ),
  caption = "Summary of body mass by penguin species and island."
)
```

## Mean Body Mass and Variation Using Standard Deviation

Mean body mass was compared across penguin species and islands. Points represent group means, with error bars showing one standard deviation around each mean.

```{r}
# Summarise body mass and pipe directly into a plot
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = mean_mass - sd_mass,
      ymax = mean_mass + sd_mass
    ),
    width = 0.2
  ) +
  labs(
    title = "Mean Body Mass Across Palmer Penguin Species and Islands",
    subtitle = "Error bars represent ±1 standard deviation",
    x = "Species",
    y = "Mean Body Mass (g)",
    colour = "Island"
  ) +
  theme_minimal()

mass_compare_plot
```

Comparing body mass across species, standard deviation shows the variation in individual body mass within each group. Larger error bars indicate larger variation in individual body size within a group, whereas smaller error bars indicate body masses to be more similar within a group. However, standard error provides a better estimate of the precision of the mean, especially when sample sizes differ.

## Mean Body Mass and Variation Using Standard Error

Mean body mass was compared across penguin species and islands. Points represent group means, with error bars showing standard error around each mean.

```{r}
# Summarise body mass and pipe directly into a plot
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = mean_mass - (sd_mass/sqrt(n)),
      ymax = mean_mass + (sd_mass/sqrt(n))
    ),
    width = 0.2
  ) +
  labs(
    title = "Mean Body Mass Across Palmer Penguin Species and Islands",
    subtitle = "Error bars represent ±1 standard error",
    x = "Species",
    y = "Mean Body Mass (g)",
    colour = "Island"
  ) +
  theme_minimal()

mass_compare_plot
```

### Ecological Interpretation

Mean body mass differed considerably among penguin species. Gentoo penguins have substantially greater body mass when compares to Adelie and Chinstrap penguins. Adelie penguins showed consistant body mass across Biscoe, Dream, and Torgersen Islands, indicating little geographic variation in body size. These patterns suggest that differences in species specific morphology have a greater influence on body size, then geographic differences within this dataset.

# Estuary Fish Survey – Data Rescue Mission

## Data Rescue and Preparation

```{r}
#| echo: false
#| message: false
#| warning: false

# Run the complete Estuary Fish Survey data-rescue workflow
source("M2_WS2_EstuaryFishSurvey.R")
```

## Clean Master Dataset

```{r}
#| echo: false

# Display a preview of the final clean master dataset
knitr::kable(
  head(master_complete, 10),
  digits = 2,
  caption = "Preview of the cleaned Estuary Fish Survey master dataset."
)
```

## Statistical Summary

```{r}
#| echo: false

knitr::kable(
  estuary_summary,
  digits = 2,
  col.names = c(
    "Scientific Name",
    "Estuary Zone",
    "Mean Catch",
    "SD Catch",
    "Mean Salinity",
    "SD Salinity"
  ),
  caption = "Mean and standard deviation of fish catch and salinity by species and estuary zone."
)
```

## Fish Abundance Along the Salinity Gradient

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Fish abundance along the estuary salinity gradient for four fish taxa."

estuary_plot
```
